Safe Transferable Regions

### 2016-07-01 paper/introduction.tex
introduction

\emph{Performance}.
When the transferred data structures are large, as is the case with many streaming big-data
analysis systems, garbage collection overhead becomes
significant~\cite{Broom:HotOS}.  Furthermore, in a distributed
dataflow system, the GC pause at one node can have a cascading adverse
effect on the performance of other nodes~\cite{Broom:HotOS,harris15}:
a GC pause at an upstream actor can block downstream actors that are waiting for
messages.  However, often much of the GC overhead results from the
collector performing avoidable or unproductive work.  For the example
in Fig.~\ref{fig:motivating-eg}, GC might repeatedly traverse the
\C{map} data structure, although its objects cannot be collected until a
suitable timing message arrives.

@@

\emph{Safety}.
These systems are designed to allow the producer and consumer to
execute in the same address space or in different address spaces
(as determined by the compiler and runtime).
We wish to ensure memory safety in the presence of such transferred data by ensuring
that there are no invalid references. A transfer operation, with no
additional checks, may cause memory safety violations, both at the
producer of the data structure, and its (possibly remote) consumer. At
the producer, any existing references into the transferred data structure become
invalid post transfer. If the data structure contains references to
objects outside (the transferred data), then such references become
invalid in the context of the consumer. Both scenarios are depicted in
Fig.~\ref{fig:transfer-violations}. Safety violations of this kind are
particularly unwelcome in the context of languages like C\# and Java
that are otherwise memory-safe. While references from within the
transferable data to outside can be disallowed in the interest of safety,
references into transferable data needs to be allowed before the data
is transferred. Allowing such references is crucial to performance,
as any non-trivial program creates temporary references to the internal
objects of a data-structure.

An important observation, in the context of processes of the kind
described above, is that the data-structures exchanged between them
can be partitioned into sets of fate-sharing objects with common
lifetimes, which makes them good candidates for a region-based memory
management discipline. A region is a block of memory that is allocated
and freed in one shot, often in constant time. A region may contain
one or more contiguous range of memory locations, and individual
objects may be dynamically allocated within the region over time,
while they are deallocated en masse when the region is freed.  Thus, a
region is a good fit for a set of fate-sharing objects.  In
Fig.~\ref{fig:motivating-eg}, the output to be constructed for each
time window $t$ (i.e., \C{map[t]}) can be a separate region that is
allocated when the first message with timestamp $t$ arrives, and
deallocated after \C{map[t]} is transferred in \C{onNotify}.

Region-based memory management, both manual as well as automatic, has
been known for a long time. Manual region-based memory management
suffers from the usual drawbacks, namely the potential for invalid
references and the consequent lack of memory safety. Automatic
region-based memory management systems guarantee memory safety, but
impose various restrictions.  MLKit, which implements the approach
pioneered by Tofte and Talpin~\cite{tofte94,tofte97}, for example,
uses lexically scoped regions.  At runtime, the set of all regions
(existing at a point in time) forms a stack. Thus, the lifetimes of
all regions must be well-nested: it is not possible to have two
regions whose lifetimes overlap, with neither one's lifetime contained
within the other.  Unfortunately, the data structures in the above
example do not satisfy this restriction (as the output messages for
multiple time windows may be simultaneously live, without any
containment relation between their lifetimes).  We refer to regions
with lexically scoped lifetimes as \emph{stack regions} and to regions
that do not have such a lexically scoped region as \emph{dynamic}
regions.

The goal of this work is a \emph{memory-safe} region-based memory
management technique that supports dynamic regions as first-class
objects. Our focus, in this paper, is on dynamic regions, which can be
safely transferred across address spaces. We refer to such dynamic
regions as \emph{transferable} regions. As with allocation and
deallocation, transferring a memory region is fast, and therefore,
transferring a data structure contained in a region is more efficient
that traversing its objects in heap, and transferring them
independently. In the \C{SelectVertex} example from
Fig.~\ref{fig:motivating-eg}, the proposed region to contain the
output for each time window $t$ must be transferable due to the
\C{transfer} operation on Line 16. As it is the case with
\C{SelectVertex}, the transferred data is no longer accessed by the
producer, so the transfer operation in our system deallocates the
region once the transfer is complete.

In this paper, we describe an approach that restores memory safety in
presence of transferable regions through a combination of a static
typing discipline and lightweight runtime checks. The cornerstone of
our approach is an \C{open} lexical block for transferable regions,
that ``opens'' a transferable region and guarantees that the region
won't be transferred/freed while it is open. Our observation is that
by nesting an MLKit-style \C{letregion} lexical block, that delimits
the lifetime of a stack region, inside an \C{open} lexical block for a
transferable region, we can guarantee that the transferable region
will remain live as long as the stack region is live. We say that the
former \emph{outlives} the later., and any references from the stack
region to the transferable region are therefore safe.  Next, we note
that by controlling the outlives relationships between various
regions, we only allow safe cross-region references, while prohibiting
unsafe ones. In the above example, an outlives relationship
\emph{from} the stack region \emph{to} the transferable region means
that the references in that direction are allowed, but not the
references in the opposite direction. In contrast, if an \C{open}
block of a transferable region $R_0$ is nested inside an \C{open}
block of another transferable region $R_1$, we do not establish any
outlives relationships, thus declaring our intention to not allow any
cross-region references between $R_0$ and $R_1$.  Finally, we observe
that outlives relationships are established based on the lexical
structure of the program, hence a static type system can enforce them
effectively. By assigning region types to objects, which capture the
regions such objects are allocated in, and by maintaining outlives
relationships between various regions, we can statically decide the
safety of all references in the program.

Indeed, the utility of our approach described above is predicated on
the assumption that we can enforce certain invariants on transferable
regions. Firstly, a transferable region cannot be transferred/freed
inside an open block of that region (i.e, while it is still open).
Secondly, a transferred/freed region cannot be opened. These are
typestate invariants on the transferable region objects, which are
hard to enforce statically due to the presence of unrestricted
aliasing. Techniques like linear types and unique pointers can be used
to restrict aliasing, but the constraints they impose are often hard
to program around. We therefore enforce typestate invariants at
runtime via lightweight checks. In particular, we define an acceptable
state transition discipline for transferable regions
(Fig.~\ref{fig:region-fsm}), and check, at runtime, whether a given
transition of a transferable region (\eg, from \emph{open} state to
\emph{freed} state) is valid or not. The check is lightweight since it
only involves checking a single tag that captures the current state.
We believe that this is a reasonable choice since regions are
coarse-grained objects manipulated infrequently, when compared to the
fine-grained objects that are present inside these regions, for which
safety is enforced statically.  An added advantage of delegating the
enforcement of typestate invariants to runtime is that our region type
system is simple, which made it possible to formulate a type inference
that completely eliminates the need to write region type annotations.
This, we believe, significantly reduces the impediment to adopt our
approach in practical setting.

@@

>>>>

\paragraph{Performance.}
When the transferred data structures are large, as is the case with many streaming big-data
analysis systems, garbage collection overhead becomes
significant~\cite{Broom:HotOS}.  Furthermore, in a distributed
dataflow system, the GC pause at one node can have a cascading adverse
effect on the performance of other nodes~\cite{Broom:HotOS,harris15}:
a GC pause at an upstream actor can block downstream actors that are waiting for
messages.  However, often much of the GC overhead results from the
collector performing avoidable or unproductive work.  For the example
in Fig.~\ref{fig:motivating-eg}, GC might repeatedly traverse the
\C{map} data structure, although its objects cannot be collected until a
suitable timing message arrives.

@@

\paragraph{Safety.}
These systems are designed to allow the producer and consumer to
execute in the same address space or in different address spaces
(as determined by the compiler and runtime).
We wish to ensure memory safety in the presence of such transferred data by ensuring
that there are no invalid references. A transfer operation, with no
additional checks, may cause memory safety violations, both at the
producer of the data structure, and its (possibly remote) consumer. At
the producer, any existing references into the transferred data structure become
invalid post transfer. If the data structure contains references to
objects outside (the transferred data), then such references become
invalid in the context of the consumer. Both scenarios are depicted in
Fig.~\ref{fig:transfer-violations}. Safety violations of this kind are
particularly unwelcome in the context of languages like C\# and Java
that are otherwise memory-safe. While references from within the
transferable data to outside can be disallowed in the interest of safety,
references into transferable data needs to be allowed before the data
is transferred. Allowing such references is crucial to performance,
as any non-trivial program creates temporary references to the internal
objects of a data-structure.

\paragraph{Safe Transferable Regions.}
In this paper, we present a language-based approach to safely and
efficiently realize such transferable data that uses regions as
the implementation technique.
%
A region is a block of memory that is allocated
and freed in one shot, often in constant time. A region may contain
one or more contiguous range of memory locations, and individual
objects may be dynamically allocated within the region over time,
while they are deallocated en masse when the region is freed.  Thus, a
region is a good fit for a transferable data-structure.  In
Fig.~\ref{fig:motivating-eg}, the output to be constructed for each
time window $t$ (i.e., \C{map[t]}) can be a separate region that is
allocated when the first message with timestamp $t$ arrives, and
deallocated after \C{map[t]} is transferred in \C{onNotify}.

The use of regions alleviates the performance concern related to
garbage collection described earlier.
However, manual region memory management introduces memory safety
issues such as dangling pointers.
We present a single type system that simultaneously addresses
this safety concern, as well as those described above related to
the use of \emph{transferable data}.

% Region-based memory management, both manual as well as automatic, has
% been known for a long time. Manual region-based memory management
% suffers from the usual drawbacks, namely the potential for invalid
% references and the consequent lack of memory safety. Automatic
% region-based memory management systems guarantee memory safety, but
% impose various restrictions.  MLKit, which implements the approach
Safe region-based memory management using types was
pioneered by Tofte and Talpin~\cite{tofte94,tofte97}, who 
use only lexically scoped regions.  At runtime, the set of all regions
(existing at a point in time) forms a stack. Thus, the lifetimes of
all regions must be well-nested: it is not possible to have two
regions whose lifetimes overlap, with neither one's lifetime contained
within the other.  Unfortunately, the data structures in the above
example do not satisfy this restriction (as the output messages for
multiple time windows may be simultaneously live, without any
containment relation between their lifetimes).  We refer to regions
with lexically scoped lifetimes as \emph{stack regions} and to regions
that do not have such a lexically scoped region as \emph{dynamic}
regions.
%
Our focus, in this paper, is on first-class \emph{memory-safe} dynamic regions
that can be safely transferred across address spaces. We refer to such dynamic
regions as \emph{transferable} regions.

Our approach is based on a variation of ideas introduced by~\cite{WW01,cyclone04}
that combine linear types and regions to suport dynamic regions.
Unlike the prior work, we ensure memory safety in the presence of
transferable regions through a combination of a static
typing discipline and lightweight runtime checks (in place of linear types).
The cornerstone of this approach is an \C{open} lexical block for transferable regions,
that ``opens'' a transferable region and guarantees that the region
won't be transferred/freed while it is open. By nesting an
MLKit-style \C{letregion} lexical block, that delimits
the lifetime of a stack region, inside an \C{open} lexical block for a
transferable region, we can guarantee that the transferable region
will remain live as long as the stack region is live. We say that the
former \emph{outlives} the later., and any references from the stack
region to the transferable region are therefore safe.  Next, we note
that by controlling the outlives relationships between various
regions, we only allow safe cross-region references, while prohibiting
unsafe ones. In the above example, an outlives relationship
\emph{from} the stack region \emph{to} the transferable region means
that the references in that direction are allowed, but not the
references in the opposite direction. In contrast, if an \C{open}
block of a transferable region $R_0$ is nested inside an \C{open}
block of another transferable region $R_1$, we do not establish any
outlives relationships, thus declaring our intention to not allow any
cross-region references between $R_0$ and $R_1$.  Finally, we observe
that outlives relationships are established based on the lexical
structure of the program, hence a static type system can enforce them
effectively. By assigning region types to objects, which capture the
regions such objects are allocated in, and by maintaining outlives
relationships between various regions, we can statically decide the
safety of all references in the program.

Ensuring memory safety using this approach requires ensuring
that the use of transferable regions satisfies certain temporal properties.
Firstly, a transferable region cannot be transferred/freed
inside an open block of that region (i.e, while it is still open).
Secondly, a transferred/freed region cannot be opened. These are
typestate invariants on the transferable region objects, which are
hard to enforce statically in the presence of unrestricted
aliasing. Techniques like linear types and unique pointers can be used
to restrict aliasing, but the constraints they impose are often hard
to program around. We therefore enforce typestate invariants at
runtime via lightweight checks. In particular, we define an acceptable
state transition discipline for transferable regions
(Fig.~\ref{fig:region-fsm}), and check, at runtime, whether a given
transition of a transferable region (\eg, from \emph{open} state to
\emph{freed} state) is valid or not. The check is lightweight since it
only involves checking a single tag that captures the current state.
We believe that this is a reasonable choice since regions are
coarse-grained objects manipulated infrequently, when compared to the
fine-grained objects that are present inside these regions, for which
safety is enforced statically.  An added advantage of delegating the
enforcement of typestate invariants to runtime is that our region type
system is simple, which made it possible to formulate a type inference
that completely eliminates the need to write region type annotations.
This, we believe, significantly reduces the impediment to adopt our
approach in practical setting.

@@

\textbf{TO DO:}

As with allocation and deallocation, transferring a memory region is fast, and therefore,
transferring a data structure contained in a region is more efficient
that traversing its objects in heap, and transferring them
independently.
In the \C{SelectVertex} example from
Fig.~\ref{fig:motivating-eg}, the proposed region to contain the
output for each time window $t$ must be transferable due to the
\C{transfer} operation on Line 16. As it is the case with
\C{SelectVertex}, the transferred data is no longer accessed by the
producer, so the transfer operation in our system deallocates the
region once the transfer is complete.

An important observation, in the context of processes of the kind
described above, is that the data-structures exchanged between them
can be partitioned into sets of fate-sharing objects with common
lifetimes, which makes them good candidates for a region-based memory
management discipline.
